Omics-based Single Step Trait Prediction

### 2017-09-15 paper/manuscript.tex
AEM suggestions up to methods-prediction

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the presence of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes~\cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
developed that allowed for a combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (that is, a
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed, \textit{i.e.}, they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

% Workings and advantages of the single step procedure.
Independently from each other,~\citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error structure.

% -   Objectives (Hybrids)
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
While single step prediction has been applied to pig hybrids~\cite{Xiang2016}
this application is not equivalent to hybrid breeding in plants.
There, the maternal and the paternal crossing partner can be selected without
constraints whereas animal breeders select only males because of the limited
fertility of the female animals.
Technological advances in plant hybrid breeding such as the rapid generation
of completely homozygous lines~\cite{Wedzony2009} now enable breeders to
produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting these hybrid candidates with low relatedness to tested hybrids using
only information from their parents is currently a hot topic in plant breeding
\cite{Technow2014,Kadam2016,Westhues2017}.
Hence, our first objective was to examine whether the single step prediction 
framework can be transfered to hybrids in plants.

% -   Objectives (quantitative predictores)
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolites and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
This situation motivated our second objective, namely exploring the utility of
single step predition with a quantitative predictor.
Yet, these technologies are not fully mature and therefore incur higher costs
to breeders than pedigree and genomic data, putting their utility into question
despite their proven usefulness.
Our third objective is therefore to evaluate the impact of equipping only a
subset of all genotypes with both, the complete and the incomplete predictor on
predictive ability.

\section{Materials and Methods}
\Genetics2level{Exp1}
To address objectives 2 and 3, a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines (described in detail by
\citeNP{Westhues2017}).
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
those primarily belonging to either cluster 1 or cluster 4 in 
Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

>>>>

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the availability of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes~\cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
developed that allow the combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step, which have a
high number of offspring, enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict breeding values
for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed and thus, they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

% Workings and advantages of the single step procedure.
Independently from each other,~\citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeN{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error structure.

% -   Objectives (Hybrids)
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
While single step prediction has been applied to pig hybrids~\cite{Xiang2016}
this application is not equivalent to hybrid breeding in plants.
There, the maternal and the paternal crossing partner can be selected without
constraints whereas animal breeders select only males because of the limited
fertility of the female animals.
Technological advances in plant hybrid breeding such as the rapid generation
of completely homozygous lines~\cite{Wedzony2009} now enable breeders to
produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting these hybrid candidates with low relatedness to tested hybrids using
only information from their parents is currently a hot topic in plant breeding
\cite{Technow2014,Kadam2016,Westhues2017}.
Hence, our first objective was to examine whether the single step prediction 
framework can be transfered to a situation where a very large number of hybrids
($\approx$ 1 million) needs to be predicted.

% -   Objectives (quantitative predictores)
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolites and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
Yet, these technologies are not fully mature and therefore incur higher costs
to breeders than pedigree and genomic data, putting their utility into question
despite their proven usefulness.
Our second objective is therefore to evaluate the impact of covering only a
subset of all genotypes with both, the complete and the incomplete predictor on
predictive ability, where for the remaining individuals only genomic data are
available.

\section{Materials and Methods}
\Genetics2level{Exp1}
To address the second objective a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17
traits~\cite{Guo2016} of which six are presented in this study
(Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were predicted with
greater precision by transcriptomic data and two further traits were predicted
with greater precision by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines described in detail by
\citeN{Westhues2017}.
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
those assigned to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.
